PACKER: An Exemplar Model of Category Generation

### 2017-01-25 Manuscripts/cogsci-2017/paper.tex
work on simulations section

As in the GCM, the similarity between two examples, $s(x_i, x_j)$, is defined as an inverse-exponential function of distance:
\begin{equation}
  s(x_i,x_j) = exp( -c \sum_{k}{|x_{ik} - x_{jk}|}w_k )
  \label{eq:similarity}
\end{equation}
where $w_k$ is a vector of attention parameters ($\sum_k{w_k} = 1$), weighting to the importance of each feature in the similarity computation, and $c$ ($<0$) is a specificity parameter controlling the spread of exemplar generalization. The attention weights are of interest in explaining individual differences in generation strategy (discussed below), but for our formal simulations they are set uniformly.

@@

In a later section, we will explore the unique predictions yielded by these design principles. First, however, we contrast PACKER with existing approaches in the field.

@@

\begin{equation}
p(z) = exp(\hat{f}_z) / \sum_z{ exp(\hat{f}_z) }
\end{equation}
%
where $\hat{f}_z$ is a function specifying each item's relative chance of being selected. As in Equation \ref{eq:packer-ss}, $\hat{f}_j$ may be set arbitrarily. However, in our simulations, $\hat{f}_z = -\gamma$ for members of contrast categories, and $\hat{f}_z = \gamma$ for members of the target category. The resulting free parameter $\gamma$ ($>0$) thus controls the relative chances that a member of the target category will be retrieved as the source. \citet{jern2013probabilistic} did not implement any mechanisms to differentially weigh retrieval probabilities, and so our model encompasses theirs as a special case when $\gamma = 0$ (uniform probabilities).

@@

@@

Because the conditions differ only in their location along the y-axis (see Figure \ref{fig:middle-bottom-conditions}), our main interest is in the generation of Beta examples above and below the contrast category. Several sample Beta categories are depicted in Figure \textbf{???????}. As is evident in Figure \textbf{???????}, we observed a great deal of individual differences in generation strategy: whereas some participants generated all four Beta examples within a narrow range, others generated Beta examples along a fairly wide range.

\begin{figure}
    \begin{center}
    \inputpgf{figs/}{distance.figs.pgf}
    \caption{Behavioral results. \textit{Top}: Frequency of exemplar generation as a function of distance from the center of the Alpha category. \textit{Bottom}: Relation between average within- and Between-category distance for every participant. }
    \label{fig:distance.figs}
    \end{center}
\end{figure}

To more specifically evaluate the key predictions made by PACKER, we determined the number of participants in each condition who placed at least one Beta example within the top `rows' of the space (one of the top two possible Y axis values), as well as the number of participants who placed at least one example in one of the bottom rows. The resulting contingency tables data are shown in Table \ref{table:subset-table}.

\begin{table}
\begin{center} 
\caption{Behavioral results.} 
\label{table:subset-table} 
\vskip 0.12in
\begin{tabular}{ l r r}
    \textbf{Middle}         & Used top rows & No top rows \\
    \hline
    Used bottom rows        &  31 & 18  \\
    No bottom rows          &  11 &  1  \\
    \\
    \textbf{Bottom}         & Used top rows & No top rows \\
    \hline
    Used bottom rows        & 22 & 8 \\
    No bottom rows          & 29 & 2 \\
\end{tabular}
\end{center} 
\end{table}

Fisher's Exact Tests reveal that a greater number of participants assigned to the Middle condition generated a Beta example in the bottom rows of the space, $p < 0.001$. Participants in the Bottom condition were more marginally more likely to generate a Beta example in the top rows of the space, $p = 0.088$. The conditions did not differ in the number of participants who placed Beta examples in the top rows \textit{and} the bottom rows, $p = 0.14$. Finally, a greater number of Middle participants generated all four Beta examples in the bottom rows of the space, $p = 0.003$, and a greater number of Bottom participants generated all four examples in the top rows of the space, $p = 0.017$.

These results are broadly supportive of PACKER's predictions, and are not predicted by existing accounts \citep[i.e.,][]{jern2013probabilistic}. However, the results described above are somewhat commonsense: they simply demonstrate that Beta examples tend to be generated in areas not occupied by the Alphas. Beyond these effects, we observed a great deal of differences in generation strategy. Figure \ref{fig:range-diff-gradient} depicts the average difference in range between the two features (e.g., horizontal -- vertical) across the participant-generated categories, with respect to the location of each category's members. These data reveal highly systematic patterns of generation: whereas many participants generated categories with more vertical range (i.e., `column'-like categories), others generated categories with horizontal range (`row'-like categories). Furthermore, these two different types of categories appear in distinct locations of the domain space. Whereas vertically aligned categories more often appear to the left or right of the Alpha class, horizontally aligned categories appear above and below the Alpha class. Thus, even beyond generating Betas in locations not occupied by the Alphas, participants appear to modify the internal structure of their categories in order to maximize distance from the Alphas.

\begin{figure}[ht!]
    \begin{center}
    \inputpgf{figs/}{range-diff-gradient.pgf}
    \caption{Generated category structure as a function of location in the domain space. Orange areas in each gradient correspond to stimuli that were commonly generated into category possessing greater vertical (Y-Axis) range. Purple areas correspond to categories possessing greater horizontal range. White areas correspond to equal range along both features (or infrequent generation).}
    \label{fig:range-diff-gradient}
    \end{center}
\end{figure}

@@

Our behavioral experiment revealed strong effects of the role of category location on generation. Not only are participants more likely to generate Betas in locations

@@

We conducted formal simulations to our data using the PACKER model described above, as well as an exemplar-based \textit{copy-and-tweak} model, and conjugate implementation of the hierarchical sampling model proposed by \citet{jern2013probabilistic}. Below, we briefly describe the formal basis of the copy-and-tweak and hierarchical sampling models.

\subsection{Model-Fitting}
In order to obtain an overall sense of each model's ability to explain our results, we fitted each model trial-wise, maximizing the log-likelihood of the its predictions against our results. For each trial, models were initialized with the participant's Alpha category (Middle or Bottom), and the configuration of the participant's Beta category during that trial. Four parameters were fitted for the PACKER model: $c$, $\phi$, $\gamma$, and $\theta$. Likewise, four parameters were fitted to the \cite{jern2013probabilistic} model: $\kappa$, $\gamma$, $\nu$, and $\theta$. Parameters were not allowed to vary between participants or conditions -- the goal was to obtain the best-fit to our entire dataset.

>>>>

As in the GCM, the similarity between two examples, $s(x_i, x_j)$, is an inverse-exponential function of distance:
\begin{equation}
  s(x_i,x_j) = exp( -c \sum_{k}{|x_{ik} - x_{jk}|}w_k )
  \label{eq:similarity}
\end{equation}
where $w_k$ is a vector of attention parameters ($\sum_k{w_k} = 1$), weighting to the importance of each feature in the similarity computation, and $c$ ($<0$) is a specificity parameter controlling the spread of exemplar generalization. The attention weights are of interest in explaining individual differences in generation strategy (discussed below), but for our formal simulations they are set uniformly.

@@

In later sections, we will explore the unique predictions yielded by these design principles. First, however, we contrast PACKER with existing approaches in the field.

@@

\begin{equation}
p(z) = exp(\hat{f}_z) / \sum_z{ exp(\hat{f}_z) }
\end{equation}
%
where $\hat{f}_z$ is a function specifying each item's relative chance of being selected. As in Equation \ref{eq:packer-ss}, $\hat{f}_j$ may be set arbitrarily. However, in our simulations, $\hat{f}_z = -\lambda$ for members of contrast categories, and $\hat{f}_z = \lambda$ for members of the target category. The resulting free parameter $\lambda$ ($>0$) thus controls the relative chances that a member of the target category will be retrieved as the source. \citet{jern2013probabilistic} did not implement any mechanisms to differentially weigh retrieval probabilities, and so our model encompasses theirs as a special case when $\lambda = 0$ (uniform probabilities).

@@

\begin{figure}[ht!]
    \begin{center}
    \inputpgf{figs/}{beta.samples.pgf}
    \caption{Sample generated categories. }
    \label{fig:beta.samples}
    \end{center}
\end{figure}

@@

Because the conditions differ only in their location along the y-axis (see Figure \ref{fig:middle-bottom-conditions}), our main interest is in the generation of Beta examples above and below the contrast category. Several sample Beta categories are depicted in Figure \ref{fig:beta.samples}. As is evident in Figure \ref{fig:beta.samples}, we observed a great deal of individual differences in generation strategy: whereas some participants generated all four Beta examples within a narrow y-axis range, others generated Beta examples along a wide range.

To more specifically evaluate the key predictions made by PACKER, we determined the number of participants in each condition who placed at least one Beta example on the top `row' of the space (the maximum possible y-axis value), as well as the number of participants who placed at least one example on the bottom row. The resulting contingencies data are shown in Table \ref{table:subset-table}. Fisher's Exact Tests on these data reveal that a greater number of participants assigned to the Middle condition generated a Beta example in the bottom row of the space, $p < 0.001$, but the conditions did not differ in use of the top of the space, $p = 0.16$. A greater number of participants in the Middle condition placed Beta examples in the top rows \textit{and} the bottom rows, $p = 0.038$.

\begin{table}
\begin{center} 
\caption{Behavioral results.} 
\label{table:subset-table} 
\vskip 0.12in
\begin{tabular}{ l r r}
    \textbf{Middle}         & Used top row & No top row \\
    \hline
    Used bottom rows       &  28 & 18  \\
    No bottom row          &  11 &  4  \\
    \\
    \textbf{Bottom}         & Used top row & No top row \\
    \hline
    Used bottom row        & 16 & 8 \\
    No bottom row          & 31 & 6 \\
\end{tabular}
\end{center} 
\end{table}

The top panel of Figure \ref{fig:distance.figs} displays the stimulus generation frequency as a function of distance from the center of the Alpha category. There was a strong preference for stimuli that are dissimilar to the members of the Alpha category: maximally distant items were by far the most frequently generated. The bottom panel of Figure \ref{fig:distance.figs} depicts for each participant the distance between exemplars, within- and between-categories. As observed by \citet{ward1994structured}, most people generated Beta categories in which members are closer to one another than they are to members of the Alpha category. We did however, observe a notable subset of individuals with greater within-class distance. Inspection of these individuals reveals that they tended to adopted a `corners' approach, in which a Beta examples were placed almost exclusively in the furthest corners of the space, thereby producing more within-class distance.

\begin{figure}
    \begin{center}
    \inputpgf{figs/}{distance.figs.pgf}
    \caption{Behavioral results. \textit{Top}: Frequency of exemplar generation as a function of distance from the center of the Alpha category. \textit{Bottom}: Relation between average within- and between-category distance for every participant. }
    \label{fig:distance.figs}
    \end{center}
\end{figure}

@@

Our results are broadly supportive of PACKER's predictions: people prefer to generate items that are maximally dissimilar from members of the Alpha category, and they tended to generate categories possessing greater within-class similarity than between-class similarity (with some notable exceptions, see Figure \ref{fig:distance.figs}). By consequence, we observed considerable differences in generation between the Middle and Bottom conditions: participants in the Bottom condition were less likely to use the bottom row of the stimulus space for generation, and participants in the Middle condition were more likely to create categories spanning the entire y-axis (utilizing the top and bottom row of the space). This latter result is especially interesting, as it shows that qualitatively different types of categories are generated, depending on only the location of the Alphas in the domain.

The results described above are somewhat commonsense: they simply demonstrate that the location (rather than distributional structure) of existing categories imposes constraints on generation because people tend to generate examples in areas not occupied by existing categories. This principle, however, is not predicted by existing models of generation -- these models are instead designed to explain distributional correspondences between generated and existing categories.

@@

We conducted formal simulations to our data using the PACKER, copy-and-tweak, and hierarchical sampling models described above. In order to obtain an overall sense of each model's ability to explain our results, we fitted each model trial-wise, using a hill-climbing algorithm to maximize the log-likelihood of the model's predictions against our results. For each trial, models were initialized with the participant's Alpha category (Middle or Bottom), and the configuration of the participant's Beta category during that trial.

Four free parameters were fitted in each model. The $c$, $\phi$, $\gamma$, and $\theta$ parameters were fitted for PACKER; $c$, $\lambda$, $\tau$, and $\theta$ were fitted for the copy-and-tweak model, and $\kappa$, $\rho$, $\nu$, and $\theta$ were fitted for the hierarchical sampling model. Note that each model possess as $\theta$ parameter fulfilling the same role (response determinism), and PACKER shares with copy-and-tweak a specificity parameter, $c$. Parameters were not allowed to vary between participants or conditions -- the goal was to obtain the best-fitting values to our entire dataset.

\begin{table}
\centering
\caption{Model-fitting results.}
\label{table:model-fits}
\begin{tabular}{ l l l}
\\
 \textbf{PACKER} & \textbf{Copy \& Tweak} & \textbf{Hierarchical} \\
 & & \textbf{Sampling} \\ \hline
 $L = -1733$ & $L = -2040$ & $L = -2055$  \\ 
 $c=0.56$ & $c=9.445$  & $\kappa<0.001$\\
$\phi=-21.2$& $\lambda=15.115$ & $\rho=1.001$ \\ 
$\gamma=18.707$  & $\tau=0.415$ & $\nu=0.451$  \\ 
 $\theta=0.166$  & $\theta=7.068$ & $\theta=6.137$  \\ 
\end{tabular}
\end{table}

Each model's best-fitting parameterization is shown in Table \ref{table:model-fits}. Overall, PACKER outperformed copy-and-tweak and the hierarchical sampling model by a considerable margin. The parameter settings associated with PACKER's best fit are exactly as expected: a strong preference for items that are similar to members of the target category but are dissimilar to members of the contrast category. A similar pattern of results was obtained when each participant's first trial was excluded (i.e., when the target category has no members).

\subsection{Category structure \& location}

\begin{figure}
    \begin{center}
    \inputpgf{figs/}{range-diff-gradient.pgf}
    \caption{Generated category structure as a function of location in the domain space. Orange areas in each gradient correspond to stimuli that were commonly generated into category possessing greater vertical (Y-Axis) range. Purple areas correspond to categories possessing greater horizontal range. White areas correspond to equal range along both features (or infrequent generation).}
    \label{fig:range-diff-gradient}
    \end{center}
\end{figure}

% Beyond these effects, we observed a great deal of differences in generation strategy. Figure \ref{fig:range-diff-gradient} depicts the average difference in range between the two features (e.g., horizontal -- vertical) across the participant-generated categories, with respect to the location of each category's members. These data reveal highly systematic patterns of generation: whereas many participants generated categories with more vertical range (i.e., `column'-like categories), others generated categories with horizontal range (`row'-like categories). Furthermore, these two different types of categories appear in distinct locations of the domain space. Whereas vertically aligned categories more often appear to the left or right of the Alpha class, horizontally aligned categories appear above and below the Alpha class. Thus, even beyond generating Betas in locations not occupied by the Alphas, participants appear to modify the internal structure of their categories in order to maximize distance from the Alphas.
